---
title: "Combining Contigs into Clones"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Combining Contigs into Clones}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 0.9em;
}
</style>
 
```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = FALSE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

suppressMessages(library(scRepertoire))
data("contig_list") 
```

There are varying definitions of clones in the literature. For the purposes of `scRepertoire`, we define a clone as cells with shared/trackable complementarity-determining region 3 (CDR3) sequences. Within this definition, one might use amino acid (`aa`) sequences of one or both chains to define a clone. Alternatively, we could use nucleotide (`nt`) or the V(D)JC genes (`genes`) to define a clone. The latter, genes, would be a more permissive definition of "clones," as multiple amino acid or nucleotide sequences can result from the same gene combination. Another option to define a clone is the use of the V(D)JC and nucleotide sequence (`strict`). `scRepertoire` allows for the use of all these definitions of clones and enables users to select both or individual chains to examine.

The first step in getting clones is to use the single-cell barcodes to organize cells into paired sequences. This is accomplished using `combineTCR()` and `combineBCR()`.

## combineTCR

The `combineTCR()` function processes a list of TCR sequencing results, consolidating them to the level of individual cell barcodes. It handles potential issues with repeated barcodes by adding prefixes from `samples` and `ID` parameters. The output includes combined reads into clone calls by nucleotide sequence (`CTnt`), amino acid sequence (`CTaa`), VDJC gene sequence (`CTgene`), or a combination of nucleotide and gene sequence (`CTstrict`).

Key Parameter(s) for `combineTCR()`

* `input.data`: A list of filtered contig annotations (e.g., filtered_contig_annotations.csv from 10x Cell Ranger) or outputs from `loadContigs()`.
* `samples`: Labels for your samples (recommended).
* `ID`: Additional sample labels (optional).
* `removeNA`: If `TRUE`, removes any cell barcode with an NA value in at least one chain (default is `FALSE`).
* `removeMulti`: If `TRUE`, removes any cell barcode with more than two immune receptor chains (default is `FALSE`).
* `filterMulti`: If `TRUE`, isolates the top two expressed chains in cell barcodes with multiple chains (default is `FALSE`).
* `filterNonproductive`: If `TRUE`, removes non-productive chains if the variable exists in the contig data (default is `TRUE`).
* `retain.sequences`: Retains full-length sequences as additional per-chain columns without altering the clone calls (default is `FALSE`). See [Retaining Full-Length Sequences](#retaining-full-length-sequences).

To combine TCR contigs from `contig_list` and apply sample prefixes:

```{r}
combined.TCR <- combineTCR(contig_list, 
                           samples = c("P17B", "P17L", "P18B", "P18L", 
                                            "P19B","P19L", "P20B", "P20L"),
                           removeNA = FALSE, 
                           removeMulti = FALSE, 
                           filterMulti = FALSE)

head(combined.TCR[[1]])
```

`combineTCR()` is the essential first step for organizing raw TCR sequencing data into a structured format for `scRepertoire` analyses. It allows for flexible handling of single and paired chains, barcode disambiguation, and initial filtering, producing a list of data frames where each row represents a single cell and its associated TCR clonotypes.

## combineBCR

The ```combineBCR()``` function is the primary tool for processing raw B cell receptor contig data into a format ready for analysis. It is analogous to ```combineTCR()``` but includes specialized logic for handling the complexities of BCRs, such as somatic hypermutation. The function consolidates contigs into a single data frame per sample, with each row representing a unique cell.

By default `(call.related.clones = TRUE)`, ```combineBCR()``` groups B cells into clones based on the similarity of their CDR3 sequences.

### How `combineBCR()` Groups Related Clones

* Internally calling ```clonalCluster()``` to build a network of related sequences.
* Using the `threshold` parameter to define connections. The `threshold` is a normalized Levenshtein distance, where a value closer to 1.0 requires higher sequence similarity. The default of 0.85 is a good starting point.
* Assigning a cluster-based ID to the CTstrict column.

Additionally, the `group.by` argument allows you to constrain the clustering analysis to only occur within distinct categories in your metadata. For example, using `group.by = "sample"` ensures that sequences from different samples are never compared or clustered together, even if they are identical.

Key Parameter(s) for `combineBCR()`

* `call.related.clones`: If `TRUE` (default), uses `clonalCluster()` to identify related clones based on sequence similarity.
* `group.by`: The column header used to group clones for clustering (if `NULL`, clusters will be calculated across all samples).
* `threshold`: The similarity threshold for `clonalCluster()` (default: 0.85).
* `dist_type`: The metric for calculating difference: "levenshtein" (default), "hamming", "nw" (Needleman-Wunsch), or "sw" (Smith-Waterman).
* `dist_mat:` The substitution matrix used if dist_type is "nw" or "sw" (e.g., "BLOSUM62", "PAM30")
* `normalize`: How to handle the threshold: "none", "length", or "maxlen".
* `chain`: The chain to use for clustering when call.related.clones = TRUE (default: `both`).
* `sequence`: The sequence type (`nt` or `aa`) for clustering (default: `nt`).
* use.V, use.J: If `TRUE`, sequences must share the same V/J gene to be clustered (default: `TRUE`)
* `cluster.method`: The clustering algorithm to apply to the edit-distanc network (default: `components`).
* `retain.sequences`: Retains full-length sequences as additional per-chain columns without altering the clone calls (default is `FALSE`). See [Retaining Full-Length Sequences](#retaining-full-length-sequences).

First, load the example BCR contig data:

```{r tidy = FALSE}
# Load example BCR contig data
BCR.contigs <- read.csv("https://www.borch.dev/uploads/contigs/b_contigs.csv")
```

Then, combine BCR contigs using the default similarity clustering:

```{r tidy = FALSE}

# Combine using the default similarity clustering
combined.BCR.clustered <- combineBCR(BCR.contigs, 
                                     samples = "Patient1", 
                                     threshold = 0.85)

# The CTstrict column contains cluster IDs (e.g., "cluster.1")
head(combined.BCR.clustered[[1]][, c("barcode", "CTstrict", "IGH", "cdr3_aa1")])
```

### Advanced: Grouping by Alignment

For more biological accuracy—specifically when analyzing amino acid sequences—you can use alignment metrics. This allows the clustering to penalize conservative amino acid changes less than radical changes.

Here we use Needleman-Wunsch alignment with the BLOSUM80 substitution matrix:

```{r}
combined.BCR.aligned <- combineBCR(BCR.contigs, 
                                   samples = "Patient1",
                                   sequence = "aa",        
                                   dist_type = "nw",      
                                   dist_mat = "BLOSUM80",  
                                   threshold = 0.85)

head(combined.BCR.aligned[[1]][, c("barcode", "CTstrict", "IGH", "cdr3_aa1")])
```


### Filtering and Cleaning Data

`combineBCR()` includes several arguments to filter and clean the contig data during processing.
* `filterNonproductive = TRUE` (Default): Removes any contigs that are not classified as productive, ensuring that only functional receptor chains are included in the analysis.
* `filterMulti = TRUE` (Default): For cells with more than one heavy or light chain detected, this automatically selects the chain with the highest UMI count (read abundance) and discards the others. This helps resolve cellular multiplets or technical artifacts.

Here is an example applying these filters (though they are on by default):

```{r}
cleaned.BCR <- combineBCR(BCR.contigs,
                          samples = "Patient1",
                          filterNonproductive = TRUE,
                          filterMulti = TRUE)

head(cleaned.BCR[[1]])
```

`combineBCR()` is designed for processing B cell repertoire data, going beyond simple contig aggregation to incorporate advanced clustering based on CDR3 sequence similarity. This enables the identification of clonally related B cells, crucial for studying B cell development, affinity maturation, and humoral immune responses. Its filtering options further ensure the quality and interpretability of the processed data.

## Retaining Full-Length Sequences

By default, `combineTCR()` and `combineBCR()` keep only the CDR3 in the `CTnt` and `CTaa` columns. Some downstream tools need the full-length receptor sequence instead. B cell lineage tracing with `dowser`, for example, reconstructs ancestral sequences from IMGT-gapped alignments, not the CDR3 alone.

To support this, `loadContigs()` standardizes a full-length `sequence` (nucleotide) and `sequence_aa` (amino acid) column for every supported format:

* AIRR-family inputs (`AIRR`, `BD`, `Dandelion`, `Immcantation`, `JSON`) carry these fields natively.
* `10X` and `MiXCR` reconstruct them by concatenating the framework and CDR regions (FWR1 through FWR4).
* CDR3-only formats (`TRUST4`, `WAT3R`, `ParseBio`) leave the columns as `NA`.
* When the source provides IMGT-gapped `sequence_alignment` and `germline_alignment` fields, those pass through as well.

The `retain.sequences` argument then carries these columns into the combined object. It is purely additive and never touches `CTaa`, `CTnt`, `CTgene`, or `CTstrict`.

* `retain.sequences = FALSE` (default): legacy output with no extra columns.
* `retain.sequences = TRUE`: retains `sequence_nt1`/`sequence_nt2` and `sequence_aa1`/`sequence_aa2` per chain.
* A character vector retains specific source columns, for example `c("sequence_alignment", "germline_alignment")` for `dowser`.

```{r eval = FALSE}
# Retain the standardized full-length sequences for both chains
combined.BCR.seq <- combineBCR(BCR.contigs,
                               samples = "Patient1",
                               retain.sequences = TRUE)

head(combined.BCR.seq[[1]][, c("barcode", "sequence_nt1", "sequence_aa1")])
```

These retained columns flow through `combineExpression()` into single-cell metadata, and `exportClones(format = "airr")` writes them into the matching AIRR rearrangement fields. For B cell lineage analysis, `exportDowser()` (or `exportClones(format = "dowser")`) reshapes a combined object built with `retain.sequences = c("sequence_alignment", "germline_alignment")` into a data frame ready for `dowser::formatClones()`. See the [Immcantation article](immcantation.html) for the full lineage workflow.

## Next Steps

- [Additional Processing Steps](Processing.html) - Add metadata, subset clones, bin by frequency, and export data.
- [Basic Clonal Visualizations](Clonal_Visualizations.html) - Visualize clonal abundance, length, and composition across samples.
- [Combining Clones and Single-Cell Objects](Attaching_SC.html) - Attach clonal data to Seurat or SCE objects with `combineExpression()`.