---
title: "TCR Specificity Grouping with immGLIPH"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{TCR Specificity Grouping with immGLIPH}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 0.9em;
}
</style>

```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = FALSE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

suppressMessages(library(scRepertoire))
suppressMessages(library(immGLIPH))
suppressMessages(library(Seurat))
suppressMessages(library(ggplot2))
suppressMessages(library(viridis))
suppressMessages(library(dplyr))

data("contig_list")
combined.TCR <- combineTCR(contig_list,
                           samples = c("P17B", "P17L", "P18B", "P18L",
                                            "P19B","P19L", "P20B", "P20L"))
```

# Getting Started

**immGLIPH** provides an R implementation of the GLIPH (Grouping of Lymphocyte Interactions by Paratope Hotspots) and GLIPH2 algorithms for clustering T cell receptors (TCRs) that are predicted to bind the same HLA-restricted peptide antigen.

The package identifies TCR specificity groups by detecting statistically enriched CDR3$\beta$ motifs (local similarity) and structurally similar CDR3$\beta$ sequences (global similarity), then clusters them into convergence groups and scores each group for biological significance.

**immGLIPH is an R implementation of existing algorithms. Users should cite the original publications:**

- **GLIPH**: Glanville, J. et al. *Identifying specificity groups in the T cell receptor repertoire.* Nature 547, 94--98 (2017). [doi:10.1038/nature22976](https://doi.org/10.1038/nature22976)

- **GLIPH2**: Huang, H. et al. *Analyzing the Mycobacterium tuberculosis immune response by T-cell receptor clustering with GLIPH2 and genome-wide antigen screening.* Nature Biotechnology 38, 1194--1202 (2020). [doi:10.1038/s41587-020-0505-4](https://doi.org/10.1038/s41587-020-0505-4)

More information is available at the [immGLIPH GitHub Repo](https://github.com/BorchLab/immGLIPH).

## Installation

```{r, eval = FALSE}
devtools::install_github("BorchLab/immGLIPH")
```

The reference repertoire data (~19 MB) is downloaded automatically the first time you run ```runGLIPH()``` and cached locally via [BiocFileCache](https://bioconductor.org/packages/BiocFileCache/).

## Integration with the scRepertoire Ecosystem

immGLIPH integrates with the [scRepertoire](https://github.com/BorchLab/scRepertoire) ecosystem through [immApex](https://github.com/BorchLab/immApex). This means ```runGLIPH()``` can directly accept:

- **Seurat** objects with TCR information
- **SingleCellExperiment** objects with TCR information
- **combineTCR()** output lists from scRepertoire
- Standard data frames or character vectors

## The Data Set

To demonstrate immGLIPH, we will use the scRepertoire example data derived from T cells of 4 patients with acute respiratory distress. The data contains paired peripheral-blood (B) and bronchoalveolar lavage (L) samples across 8 distinct runs from the 10x Cell Ranger pipeline.

# Loading Libraries and Data

```{r}
scRep_example <- readRDS("scRep_example_full.rds")

scRep_example <- combineExpression(combined.TCR,
                                   scRep_example,
                                   cloneCall="aa",
                                   group.by = "sample",
                                   proportion = TRUE)

#Filtering for single-cells with TCRs
scRep_example <- subset(scRep_example,
                        cells = colnames(scRep_example)[!is.na(scRep_example$CTaa)])
```

# Running GLIPH

## Input Data Formats

```runGLIPH()``` accepts multiple input formats. The most direct integration with scRepertoire is passing the ```combineTCR()``` output or a Seurat object:

```{r, eval = FALSE}
# Option 1: Pass combineTCR() output directly
results <- runGLIPH(combined.TCR, method = "gliph2")

# Option 2: Pass a Seurat object with TCR metadata
results <- runGLIPH(scRep_example, method = "gliph2", chains = "TRB")
```

## Running GLIPH2 (Default)

By default ```runGLIPH()``` downloads a reference repertoire via BiocFileCache. For this vignette, we will extract TCR data and supply a custom reference data frame directly using ```refdb_beta``` to demonstrate the workflow:

```{r}
# Extract TRB CDR3 and V-gene information from scRepertoire data
tcr_df <- exportClones(scRep_example,
                       write.file = FALSE,
                       format = "immGLIPH")
tcr_df <- na.omit(tcr_df)

# Use a subset as a small reference for demonstration
ref_df <- tcr_df[, c("CDR3b", "TRBV")]

res_gliph2 <- runGLIPH(
  cdr3_sequences = tcr_df,
  method         = "gliph2",
  refdb_beta     = ref_df,
  sim_depth      = 100,
  n_cores        = 1
)
```

## Running GLIPH1

The original GLIPH algorithm uses repeated random sampling for local similarity detection and Hamming distance for global similarity:

```{r}
res_gliph1 <- runGLIPH(
  cdr3_sequences = tcr_df,
  method         = "gliph1",
  refdb_beta     = ref_df,
  sim_depth      = 100,
  n_cores        = 1
)
```

## Method Presets

The ```method``` parameter configures a coordinated set of algorithm choices:

| Setting | `"gliph1"` | `"gliph2"` | `"custom"` |
|:--------|:-----------|:-----------|:-----------|
| Local method | Repeated random sampling | Fisher's exact test | User-defined |
| Global method | Hamming distance cutoff | Struct-based + Fisher | User-defined |
| Clustering | Connected components | Per-motif isolated | User-defined |
| Scoring | GLIPH1 formula | GLIPH2 formula | User-defined |

# Understanding the Output

The output is a list with several components:

```{r}
names(res_gliph2)
```

## Cluster Properties

The ```cluster_properties``` data frame summarizes each convergence group with enrichment scores:

```{r}
head(res_gliph2$cluster_properties)
```

## Cluster Membership

The ```cluster_list``` is a named list where each element contains the member TCRs of a convergence group:

```{r}
length(res_gliph2$cluster_list)

if (length(res_gliph2$cluster_list) > 0) {
  head(res_gliph2$cluster_list[[1]])
}
```

## Motif Enrichment

The ```motif_enrichment``` element contains the locally enriched motifs:

```{r}
if (!is.null(res_gliph2$motif_enrichment$selected_motifs)) {
  head(res_gliph2$motif_enrichment$selected_motifs)
}
```

## Network Edges

The ```connections``` data frame contains the edge list representing the TCR similarity network:

```{r}
if (!is.null(res_gliph2$connections)) {
  head(res_gliph2$connections)
}
```

# Motif Discovery with findMotifs()

The ```findMotifs()``` function searches for continuous and discontinuous k-mer motifs in a set of CDR3 sequences. It is used internally by ```runGLIPH()``` but can also be called independently:

```{r}
sample_seqs <- as.character(tcr_df$CDR3b[seq_len(200)])

# Find all 3-mers appearing at least 5 times
motifs <- findMotifs(seqs = sample_seqs, q = 3, kmer_mindepth = 5)
head(motifs[order(motifs$V1, decreasing = TRUE), ])
```

Including discontinuous motifs (e.g., ```C.S``` where ```.``` is any amino acid):

```{r}
disc_motifs <- findMotifs(
  seqs         = sample_seqs,
  q            = 2,
  kmer_mindepth = 5,
  discontinuous = TRUE
)
disc_only <- disc_motifs[grep("\\.", disc_motifs$motif), ]
head(disc_only[order(disc_only$V1, decreasing = TRUE), ])
```

# Customizing the Analysis

## Using method = "custom"

The ```"custom"``` method allows independent control over each algorithmic component:

```{r, eval = FALSE}
res_custom <- runGLIPH(
  cdr3_sequences  = tcr_df,
  method          = "custom",
  local_method    = "fisher",
  global_method   = "cutoff",
  clustering_method = "GLIPH1.0",
  scoring_method  = "GLIPH2.0",
  sim_depth       = 100,
  n_cores         = 1
)
```

## Adjusting Significance Thresholds

For the Fisher-based local method (GLIPH2), you can adjust:

- **```lcminp```**: Maximum p-value for motif enrichment (default 0.01)
- **```lcminove```**: Minimum fold-enrichment per motif length (default ```c(1000, 100, 10)``` for 2-mers, 3-mers, 4-mers)
- **```kmer_mindepth```**: Minimum motif observations in the sample (default 3)

```{r, eval = FALSE}
res_strict <- runGLIPH(
  cdr3_sequences = tcr_df,
  method         = "gliph2",
  lcminp         = 0.001,
  lcminove       = c(10000, 1000, 100),
  sim_depth      = 100,
  n_cores        = 1
)
```

## Choosing a Reference Database

immGLIPH ships with reference repertoires for human and mouse from the original GLIPH publications:

| Name | Species | Version | Subset |
|:-----|:--------|:--------|:-------|
| `"human_v2.0_CD48"` | Human | v2.0 | CD4+CD8 (default) |
| `"human_v2.0_CD4"` | Human | v2.0 | CD4 |
| `"human_v2.0_CD8"` | Human | v2.0 | CD8 |
| `"human_v1.0_CD48"` | Human | v1.0 | CD4+CD8 |
| `"mouse_v1.0_CD48"` | Mouse | v1.0 | CD4+CD8 |

# Network Visualization

The ```plotNetwork()``` function creates an interactive network visualization of the convergence groups using the visNetwork package:

```{r, eval = FALSE}
plotNetwork(
  clustering_output = res_gliph2,
  color_info        = "total.score",
  cluster_min_size  = 3,
  n_cores           = 1
)
```

You can also color nodes by donor/patient:

```{r, eval = FALSE}
plotNetwork(
  clustering_output = res_gliph2,
  color_info        = "patient",
  cluster_min_size  = 3,
  n_cores           = 1
)
```

# De Novo TCR Generation

The ```deNovoTCRs()``` function generates artificial CDR3$\beta$ sequences that resemble the positional amino acid composition of a given convergence group. This can be used to predict novel TCR sequences with similar binding characteristics:

```{r, eval = FALSE}
de_novo <- deNovoTCRs(
  convergence_group_tag = res_gliph2$cluster_properties$tag[1],
  clustering_output     = res_gliph2,
  sims                  = 10000,
  num_tops              = 100,
  make_figure           = TRUE,
  n_cores               = 1
)

head(de_novo$de_novo_sequences)
```

# Performance

## Accelerated Computation with immApex

When [immApex](https://github.com/BorchLab/immApex) is installed, immGLIPH automatically uses its C++-accelerated backends for two computationally intensive steps:

1. **Motif enumeration** (```findMotifs()```): Uses ```immApex::calculateMotif()``` with OpenMP multithreading.
2. **Global Hamming distance network** (GLIPH1 method): Uses ```immApex::buildNetwork()``` to compute pairwise distances in a single C++ call.

If immApex is not installed, immGLIPH falls back to the original pure-R implementations transparently.

```{r, eval = FALSE}
devtools::install_github("BorchLab/immApex")
```

# Tips and Best Practices

1. **Start with GLIPH2**: The Fisher-based approach is generally more statistically rigorous than repeated random sampling.
2. **Sample size matters**: GLIPH works best with >200 unique CDR3$\beta$ sequences.
3. **Include V-gene information**: When available, TRBV data improves both global similarity detection and scoring accuracy.
4. **Adjust ```sim_depth```**: For publication-quality results, use ```sim_depth >= 1000```. For exploratory analysis, ```sim_depth = 100``` is faster.
5. **Parallelization**: For large datasets (>5,000 sequences), set ```n_cores > 1```.
6. **Install immApex**: For best performance, install immApex to enable C++-accelerated motif enumeration and network construction.
