---
title: "Immune Reference Data with immReferent"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Immune Reference Data with immReferent}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 0.9em;
}
</style>

```{r, include = FALSE}
all_times <- list()
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = FALSE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

has_immReferent <- requireNamespace("immReferent", quietly = TRUE)

if (has_immReferent) {
  library(immReferent)
  imgt_ok  <- try(is_imgt_available(),  silent = TRUE)
  ogrdb_ok <- try(is_ogrdb_available(), silent = TRUE)
  imgt_ok  <- if (inherits(imgt_ok,  "try-error")) FALSE else isTRUE(imgt_ok)
  ogrdb_ok <- if (inherits(ogrdb_ok, "try-error")) FALSE else isTRUE(ogrdb_ok)
} else {
  imgt_ok  <- FALSE
  ogrdb_ok <- FALSE
}

can_run <- has_immReferent & imgt_ok

set.seed(42)
```

## Introduction

The [immReferent](https://github.com/BorchLab/immReferent) package provides a centralized interface for downloading, managing, and loading immune repertoire and HLA reference sequences from [IMGT](https://www.imgt.org/), [IPD-IMGT/HLA](https://www.ebi.ac.uk/ipd/imgt/hla/), and [OGRDB](https://ogrdb.airr-community.org/). It serves as a core dependency for immunogenomics packages, ensuring reliable and high-quality sequence access with local caching for reproducibility.

In the context of scRepertoire, immReferent is useful for:

* Obtaining the reference gene sequences used in TCR/BCR analysis
* Exporting formatted references for tools like MiXCR, TRUST4, Cell Ranger, and IgBLAST
* Ensuring offline reproducibility through its caching system

## Installation

```{r eval = FALSE}
devtools::install_github("BorchLab/immReferent")
```

Or via Bioconductor:

```{r eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("immReferent")
```

## Downloading Reference Sequences

```{r setup, eval = can_run}
library(immReferent)
```

### HLA Sequences (IPD-IMGT/HLA)

The IPD-IMGT/HLA database provides reference sequences for the Human Leukocyte Antigen (HLA) system.

```{r get_hla, eval = can_run}
hla_prot <- getIMGT(gene = "HLA",
                    type = "PROT")

print(hla_prot)
cat("Number of sequences:", length(hla_prot), "\n")
```

### TCR/BCR Sequences (IMGT)

For T-cell receptor (TCR) and B-cell receptor (BCR) genes, specify the species and gene or gene family.

```{r get_ighv, eval = can_run}
# Download human IGHV nucleotide sequences
ighv_nuc <- getIMGT(species = "human",
                    gene = "IGHV",
                    type = "NUC")
print(ighv_nuc)
```

You can also download entire families of genes at once:

```{r get_trb, eval = can_run}
# Download all mouse TRB genes (V, D, J, and C)
trb_mouse <- getIMGT(species = "mouse",
                     gene = "TRB",
                     type = "NUC")
print(trb_mouse)
```

### Germline Sets from OGRDB (AIRR)

OGRDB provides AIRR-compliant germline sets for immunoglobulin loci.

```{r ogrdb_igh, eval = has_immReferent & ogrdb_ok}
# Human IGH nucleotide sequences (gapped FASTA)
igh_ogrdb <- getOGRDB(
  species = "human",
  locus   = "IGH",
  type    = "NUC",
  format  = "FASTA_GAPPED"
)
igh_ogrdb
```

## Exporting for External Tools

immReferent can export reference sequences formatted for popular analysis tools.

```{r export_setup, eval = can_run}
igh_seqs <- getIMGT(species = "human",
                    gene = "IGH",
                    type = "NUC",
                    suppressMessages = TRUE)
```

### MiXCR

```{r export_mixcr, eval = can_run}
mixcr_dir <- tempdir()
mixcr_files <- exportMiXCR(igh_seqs,
                           mixcr_dir,
                           chain = "IGH")
print(mixcr_files)
```

### TRUST4

```{r export_trust4, eval = can_run}
trust4_file <- tempfile(fileext = ".fa")
exportTRUST4(igh_seqs, trust4_file)
cat(head(readLines(trust4_file), 4), sep = "\n")
```

### Cell Ranger VDJ

```{r export_cellranger, eval = can_run}
cellranger_file <- tempfile(fileext = ".fa")
exportCellRanger(igh_seqs, cellranger_file)
cat(head(readLines(cellranger_file), 4), sep = "\n")
```

### IgBLAST

```{r export_igblast, eval = can_run}
igblast_dir <- tempdir()
igblast_files <- exportIgBLAST(igh_seqs, igblast_dir,
                               organism = "human",
                               receptor_type = "ig")
print(igblast_files)
```

## Caching and Offline Usage

immReferent automatically caches all downloaded data locally. On subsequent requests, the cached copy is loaded without network access.

```{r list_cache, eval = can_run}
# List all cached IMGT files
listIMGT()
```

To load strictly from cache (no downloads):

```{r load_cache, eval = can_run}
cached <- loadIMGT(species = "human",
                   gene = "IGHV",
                   type = "NUC")
```

To force a re-download when the online data has been updated:

```{r refresh, eval = can_run}
fresh <- refreshIMGT(species = "human",
                     gene = "IGHV",
                     type = "NUC")
```

You can change the cache location for a session or permanently via `.Rprofile`:

```{r eval = FALSE}
options(immReferent.cache = "/path/to/shared/cache")
```

## Session Info

```{r}
sessionInfo()
```
