---
title: "Additional Processing Steps"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Additional Processing Steps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 0.9em;
}
</style>
 
```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = FALSE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

suppressMessages(library(scRepertoire))
data("contig_list") 
combined.TCR <- combineTCR(contig_list, 
                           samples = c("P17B", "P17L", "P18B", "P18L", 
                                       "P19B","P19L", "P20B", "P20L"))
```

## addVariable: Adding Variables for Plotting

What if there are more variables to add than just sample and ID? We can add them by using the `addVariable()` function. For each element, the function will add a column (labeled by `variable.name`) with the `variable`. The length of the `variables` parameter needs to match the length of the combined object.

Key Parameter(s) for `addVariable()`

* `variable.name`: A character string that defines the new variable to add (e.g., "Type", "Treatment").
* `variables`: A character vector defining the desired column value for each list element. Its length must match the number of elements in the input.data list.

As an example, here we add the Type in which the samples were processed and sequenced to the `combined.TCR` object:

```{r}
combined.TCR <- addVariable(combined.TCR, 
                            variable.name = "Type", 
                            variables = rep(c("B", "L"), 4))

head(combined.TCR[[1]])
```

## subsetClones: Filter Out Clonal Information

Likewise, we can remove specific list elements after `combineTCR()` or `combineBCR()` using the `subsetClones()` function. In order to subset, we need to identify the column header we would like to use for subsetting (`name`) and the specific values to include (`variables`).

Key Parameter(s) for `subsetClones()`

* `name`: The column header/name in the metadata of input.data to use for subsetting (e.g., "sample", "Type").
* `variables`: A character vector of the specific values within the chosen name column to retain in the subsetted data.

Below, we isolate just the two sequencing results from "P18L" and "P18B" samples:

```{r}
subset1 <- subsetClones(combined.TCR, 
                        name = "sample", 
                        variables = c("P18L", "P18B"))

head(subset1[[1]][,1:4])
```

Alternatively, we can also just select the list elements after `combineTCR()` or `combineBCR()`.

```{r}
subset2 <- combined.TCR[c(3,4)]
head(subset2[[1]][,1:4])
```

## exportClones: Save Clonal Data  

After assigning the clone by barcode, we can export the clonal information using `exportClones()` to save for later use or to integrate with other bioinformatics pipelines. This function supports various output formats tailored for different analytical needs.

Key Parameter(s) for `exportClones()`
* `format`: The desired output format for the clonal data.
  * `airr`: Exports data in an Adaptive Immune Receptor Repertoire (AIRR) Community-compliant format, with each row representing a single receptor chain.
   * `immunarch`: Exports a list containing a data frame and metadata formatted for use with the `immunarch` package. 
  * `paired`: Exports a data frame with paired chain information (amino acid, nucleotide, genes) per barcode. This is the default.
  * `TCRMatch`: Exports a data frame specifically for the TCRMatch algorithm, containing TRB chain amino acid sequence and clonal frequency.
  * `tcrpheno`: Exports a data frame compatible with the `tcrpheno` pipeline, with TRA and TRB chains in separate columns.
* `write.file`: If `TRUE` (default), saves the output to a CSV file. If `FALSE`, returns the data frame or list to the R environment.
* `dir`: The directory where the output file will be saved. Defaults to the current working directory.
* `file.name`: The name of the CSV file to be saved.

To export the combined clonotypes as a `paired` data frame and save it to a specified directory:

```{r, eval = FALSE}
exportClones(combined, 
             write.file = TRUE,
             dir = "~/Documents/MyExperiment/Sample1/"
             file.name = "clones.csv")
```

To return an `immunarch`-formatted data frame directly to your R environment without saving a file:

```{r}
immunarch <- exportClones(combined.TCR, 
                          format = "immunarch", 
                          write.file = FALSE)
head(immunarch[[1]][[1]])
```

## clonalBin: Bin Clones by Frequency or Proportion

The `clonalBin()` function adds a clonal grouping variable (`cloneSize`) to the output of `combineTCR()`, `combineBCR()`, or `combineExpression()`. This function calculates the clonal frequency and proportion, then bins clones into categories based on customizable thresholds. This is useful for categorizing clones prior to downstream analysis or visualization, without needing to attach the data to a single-cell object.

Key Parameter(s) for `clonalBin()`

* `clone.call`: Defines the clonal sequence grouping (`gene`, `nt`, `aa`, or `strict`).
* `chain`: The TCR/BCR chain to use (`both`, `TRA`, `TRB`, `TRG`, `TRD`, `IGH`, `IGL`).
* `group.by`: A column header in the metadata to group the analysis by (e.g., "sample", "Type").
* `proportion`: Whether to use proportion (`TRUE`) or total frequency (`FALSE`) for binning.
* `clone.size`: The bins for grouping based on proportion or frequency.

### Basic Usage with Proportion

By default, `clonalBin()` uses proportion-based binning with the following default bins: `c(Rare = 1e-4, Small = 0.001, Medium = 0.01, Large = 0.1, Hyperexpanded = 1)`.

```{r}
combined.TCR <- clonalBin(combined.TCR, clone.call = "strict")

# Check the cloneSize column
head(combined.TCR[[1]][, c("barcode", "CTstrict", "clonalProportion", "clonalFrequency", "cloneSize")])

# View the distribution of clone sizes
table(combined.TCR[[1]]$cloneSize)
```

### Using Frequency-Based Binning

When using frequency-based binning (`proportion = FALSE`), the `clone.size` values must be integers. If the maximum frequency exceeds the upper bin limit, it will be automatically adjusted.

```{r}
combined.TCR.freq <- combineTCR(contig_list,
                                samples = c("P17B", "P17L", "P18B", "P18L",
                                            "P19B","P19L", "P20B", "P20L"))

combined.TCR.freq <- clonalBin(combined.TCR.freq,
                               clone.call = "strict",
                               proportion = FALSE,
                               clone.size = c(Rare = 1, Small = 5, Medium = 20,
                                              Large = 100, Hyperexpanded = 500))

# Check the frequency-based binning
head(combined.TCR.freq[[1]][, c("barcode", "clonalFrequency", "cloneSize")])
```

### Grouping by a Variable

You can calculate clonal frequency and proportion across samples grouped by a variable (e.g., "Type" or "sample"). This is particularly useful when you want consistent binning across experimental conditions.

```{r}
combined.TCR.grouped <- combineTCR(contig_list,
                                   samples = c("P17B", "P17L", "P18B", "P18L",
                                               "P19B","P19L", "P20B", "P20L"))

combined.TCR.grouped <- addVariable(combined.TCR.grouped,
                                    variable.name = "Type",
                                    variables = rep(c("B", "L"), 4))

combined.TCR.grouped <- clonalBin(combined.TCR.grouped,
                                  clone.call = "strict",
                                  group.by = "Type")

# Check grouped binning
head(combined.TCR.grouped[[1]][, c("barcode", "Type", "clonalProportion", "cloneSize")])
```

## annotateInvariant

The `annotateInvariant()` function enables the identification of mucosal-associated invariant T (`MAIT`) cells and invariant natural killer T (`iNKT`) cells in single-cell sequencing datasets. These specialized T-cell subsets are defined by their characteristic TCR usage, making them distinguishable within single-cell immune profiling data. The function extracts TCR chain information from the provided single-cell dataset and evaluates it against known invariant TCR criteria for either MAIT or iNKT cells. Each cell is assigned a score indicating the presence (1) or absence (0) of the specified invariant T-cell population.

Key Parameter(s) for `annotateInavriant()`

* `type`: Character string specifying the type of invariant T cell to annotate (`MAIT` or `iNKT`).
* `species`: Character string specifying the species (`mouse` or `human`).

```{r, eval = FALSE}
combined <- annotateInvariant(combined, 
                              type = "MAIT", 
                              species = "human")

combined <- annotateInvariant(combined,
                              type = "iNKT",
                              species = "human")
```

## Next Steps

- [Basic Clonal Visualizations](Clonal_Visualizations.html) - Visualize clonal abundance, length, and composition across samples.
- [Combining Clones and Single-Cell Objects](Attaching_SC.html) - Attach clonal data to Seurat or SCE objects with `combineExpression()`.
- [Summarizing Repertoires](Repertoire_Summary.html) - Analyze gene usage, amino acid properties, and k-mer motifs.